---
title: "Reproducible Research: Peer Assessment 1"
author: Dexter Wang
date: "08/05/2016"
output: 
  html_document:
    keep_md: yes
---


## Loading and preprocessing the data

```{r echo=TRUE}

setwd("C:/D/R/Output Repo/Assignments Temp Repo/RepData_PeerAssessment1")

unzip("activity.zip")

data <- read.csv("./activity.csv",sep = ",",header= TRUE)

data$date <- as.Date(as.character(data$date),"%Y-%m-%d") 


str(data)

head(data)


```

####There are `r nrow(data)` observations in the dataset, dated from `r min(data$date)` to `r max(data$date)`. Each day has `r table(data$date)[[1]]` observations. Each records the number of steps taken in the sequential of every 5 minutes interval. 

####There is `r paste("",round(mean(is.na(data$steps))*100),"%") ` data missing in "steps" column


## What is mean total number of steps taken per day?

```{r message=FALSE,warning=FALSE}

if (!require("ggplot2")) {
  install.packages("ggplot2")
}

library(ggplot2)


Daily_Steps <- with(data,tapply(steps,date,sum))

Daily_Steps <- as.data.frame(Daily_Steps)

meanValue <- mean(Daily_Steps$Daily_Steps,na.rm=TRUE)

medianValue <- median(Daily_Steps$Daily_Steps,na.rm=TRUE)

g <- ggplot(Daily_Steps,aes(x=Daily_Steps))+geom_histogram(bins = 10)

g+geom_vline(aes(xintercept=meanValue),size=1.5,linetype="dashed",colour="red" )+annotate("text", x=meanValue-800,y=7, label= paste("mean = ",round(meanValue)),angle=90,size=8,colour="red")+labs(x="total steps per day",y="",title="histogram of the total number of steps taken each day")
```

####The mean of the total number of steps taken per day is `r as.integer(round(meanValue))`

####The median of the total number of steps taken per day is `r medianValue`

####The NA values in steps column are removed from the calculation


## What is the average daily activity pattern?

``` {r message=FALSE,warning=FALSE}
#prepare data

Steps_time_interval <- with(data,tapply(steps,interval,mean,na.rm=TRUE))

Steps_time_interval <- as.data.frame(Steps_time_interval)

Steps_time_interval$interval <- as.numeric(row.names(Steps_time_interval))

names(Steps_time_interval) <- c("steps","interval")

max_steps <- max(Steps_time_interval$steps)
time_max_steps <- Steps_time_interval[Steps_time_interval$steps == max_steps,]$interval

#prepare plot

ann_text <- paste("maximum ",round(max_steps)," steps at ",round(time_max_steps/100),":",time_max_steps-round(time_max_steps/100)*100," am")

# x axis time sequence 
x_axis <- seq(0,2355,100)

time_axis <- seq(as.POSIXct("1910/1/1"), as.POSIXct("1910/1/1 23:55"), "hours")

g <- ggplot(Steps_time_interval,aes(x=interval,y=steps,group=1))+geom_line()

g + scale_x_continuous(breaks=x_axis,labels=format(time_axis,"%H:%M"))+theme(axis.text.x = element_text(angle=45))+annotate("text",x=time_max_steps,y=max_steps+10,label=ann_text,size=6)+labs(x="time of day",y="average number of steps",title="daily activity pattern")


```

#### The maximum number of steps (in 5 minutes interval) is `r round(max_steps)` in average taken `r paste("at ",round(time_max_steps/100),":",time_max_steps-round(time_max_steps/100)*100," am")`



## Imputing missing values

####There are `r nrow(data)` observations in the dataset

####There are `r sum(is.na(data$steps)) ` rows with missing "steps" value


#### replace missing data using average steps by each time interval
```{r message=FALSE,warning=FALSE}

# replace missing data using average steps by time interval 
data_NonNA <- merge(data,Steps_time_interval,by.x="interval",by.y="interval")

data_NonNA<-data_NonNA[order(data_NonNA$date,data_NonNA$interval),] 

data_NonNA[is.na(data_NonNA$steps.x),]$steps.x <- data_NonNA[is.na(data_NonNA$steps.x),]$steps.y

data_NonNA$steps.y <-NULL

names(data_NonNA)[2]<- "steps"

str(data_NonNA)

head(data_NonNA)


Daily_Steps_NonNA <- with(data_NonNA,tapply(steps,date,sum))

Daily_Steps_NonNA <- as.data.frame(Daily_Steps_NonNA)

meanValue <- mean(Daily_Steps_NonNA$Daily_Steps_NonNA)

medianValue <- median(Daily_Steps_NonNA$Daily_Steps_NonNA)


g <- ggplot(Daily_Steps_NonNA,aes(x=Daily_Steps_NonNA))+geom_histogram(bins = 12)

g+geom_vline(aes(xintercept=meanValue),size=1.5,linetype="dashed",colour="blue" )+annotate("text", x=meanValue-800,y=9, label= paste("mean / median = ",round(meanValue)),angle=90,size=8,colour="blue")+labs(x="total steps per day",y="",title="histogram of the total number of steps taken each day \n with missing step values replaced by \n average steps taken in the corresponding time interval")


```

####The mean of the total number of steps taken per day is `r as.integer(round(meanValue))`

####The median of the total number of steps taken per day is `r as.integer(round(medianValue))`

####It shows that the mean and median value have not changed after replacing the missing values using the average steps by time interval



## Are there differences in activity patterns between weekdays and weekends?

```{r message=FALSE,warning=FALSE}

if (!require("reshape")) {
  install.packages("reshape")
}

library(reshape)

# prepare data 

data_NonNA$weekday <- weekdays(data_NonNA$date)

data_NonNA$is.weekend <-data_NonNA$weekday %in% c("Saturday","Sunday") 

data_NonNA$is.weekend <- as.character(data_NonNA$is.weekend)

data_NonNA[data_NonNA$is.weekend == "TRUE","is.weekend"] <- "Weekend"

data_NonNA[data_NonNA$is.weekend == "FALSE","is.weekend"] <- "Weekday"


Steps_time_interval_NonNA <- with(data_NonNA,tapply(steps,list(interval,is.weekend),mean))

Steps_time_interval_NonNA <- as.data.frame(Steps_time_interval_NonNA)

Steps_time_interval_NonNA$interval <- as.numeric(row.names(Steps_time_interval_NonNA))

names(Steps_time_interval_NonNA) <- c("weekday","weekend","interval")

Steps_time_interval_NonNA <- melt(Steps_time_interval_NonNA ,  id.vars = "interval", variable.name = 'steps')

names(Steps_time_interval_NonNA)<-c("interval","is.weekday","average.steps")

str(Steps_time_interval_NonNA)

head(Steps_time_interval_NonNA,10)


#prepare plot

# x axis time sequence 
x_axis <- seq(0,2355,100)

time_axis <- seq(as.POSIXct("1910/1/1"), as.POSIXct("1910/1/1 23:55"), "hours")

g <- ggplot(Steps_time_interval_NonNA, aes(x=interval,y=average.steps))+geom_line(aes(group=is.weekday,colour=is.weekday),size=2,alpha=0.8)

g + scale_x_continuous(breaks=x_axis,labels=format(time_axis,"%H:%M"))+theme(axis.text.x = element_text(angle=45))+labs(x="time of day",y="average number of steps",title="daily activity pattern in weekday/weekends")

```

####I prefer to draw the lines together in a single plot, but the requirment seems need the lines to be separate. So here we go...

```{r}


g <- ggplot(Steps_time_interval_NonNA, aes(x=interval,y=average.steps))+geom_line(aes(group=is.weekday,colour=is.weekday),size=2,alpha=0.8)+facet_grid(is.weekday~.)



g + scale_x_continuous(breaks=x_axis,labels=format(time_axis,"%H:%M"))+theme(axis.text.x = element_text(angle=45))+labs(x="time of day",y="average number of steps",title="daily activity pattern in weekday/weekends")



```


####It appears that people take more steps in the afternoon during weekends compare to that in weekdays.In weekdays, people are more active in early morning before 9:00am.

##Thanks for viewing this : )